Peaks Over Threshold for Bursty Time Series

### 2018-04-21 article_springer/article_springer.Rmd
minor changes

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
For all statistical computations we have used R [@R], and all code, data, as 
well as the source code for this manuscript has been organized into an R 
software package, available at <https://github.com/UNSW-MATH/CTRM>.

@@

```{r thresholdedBursty, message=FALSE, fig.height=7, fig.cap="\\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line)."}
#rpareto <- function(n,tail){
#          if(tail<=0) stop("tail must be positive")
#          rp <- (runif(n)/gamma(1-tail))^(-1/tail)
#          rp}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
#times<-cumsum(rpareto(n,tail))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctrm <- ctrm(data.frame(times, magnitudes)) 
par(mfrow = c(2,1))
plot(sim_ctrm, p = 0.01, main = "Simulated MRP") 
flares %>% ctrm() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") 
```

@@

\noindent For a scale parameter $\sigma > 0$, we write 
${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. 
The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, from the point of view of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11.

@@

```{r fig:QQ-est, fig.cap="QQ-estimator plots for various datasets.\\label{fig:QQ-est}", fig.height=7, fig.width=7}
par(mfrow = c(2,2))
flares_exponent <- flares %>% ctrm() %>% thin(k=500) %>% qqestplot(static=TRUE, main = "QQ-static, flare magnitudes")
ML_exponent <- MittagLeffleR::rml(500, 0.8) %>% qqestplot(static=TRUE, main = "QQ-static, Mittag-Leffler data")
ML_exponent2 <- MittagLeffleR::rml(500, 0.8) %>% qqestplot(static=FALSE, top_k = 15, conf.int = TRUE, main = "lala")
exp_exponent <- rexp(500) %>% qqestplot(static=TRUE, main = "QQ-static, Exponential data")
```

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis). The reciprocal slope is returned as the estimate of the tail 
exponent.
For instance, the left panel in Figure \ref{fig:QQ-plots} shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill estimator is. 
The top-right panel of Figure \ref{fig:QQ-plots}, for instance, shows the 
QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). Hence a cutoff needs to be introduced, 
above which the tail exponent is fitted: The dynamic QQ-estimator 
plots the static QQ-estimates (y-axis) for different cutoffs (x-axis)
(see bottom-left panel of Figure \ref{fig:QQ-est}). There is no apparent interval of stability, 
which could be taken to indicate a reasonable estimate for the true 
tail exponent $0.8$. 
Even worse, from a QQ-estimator point of view, there is no striking 
difference between the infinite mean ${\rm ML}(0.8,1)$ distribution and 
the light-tailed ${\rm Exp}(1)$ distribution (compare the right two panels). 
In particular, if only few (e.g. $<100$) data are available, 
then the static QQ-estimator plots look interchangeable for the two 
distributions.

@@

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly 
(see Figure \ref{fig:QQ-plots}). 
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in the R package `MittagLeffleR` 
[@MittagLeffleR]. 
Since the exponential distribution is nested in the Mittag-Leffler 
family of distributions, a standard likelihood ratio test can be 
performed, with the exponential distribution as a null model against 
a Mittag-Leffler distribution as the alternative. 
As an example, the threshold crossing times for the solar flare dataset
yield a difference in deviance of 
$\approx$ `r round(2 * (mlmle_out$loglik - mlmle_out$null_loglik))`, 
which evaluates to a strongly significant $\chi^2_1$ test statistic 
(p-value = $10^{`r round(log10(mlmle_out$p_value))`}$) for the null
hypothesis $\beta = 1$.

```{r QQ-plots, fig.cap="QQ Plots for Mittag-Leffler data (left; $\\beta = 0.8$) and threshold crossing times / magnitudes for the solar flare data (middle / right; threshold is set at the 500th order statistic, $\\beta = 0.83$). Threshold crossing times appear to fit the Mittag-Leffler distribution well, whereas the Magnitudes appear to fit a Pareto distribution better (\\ref{fig:QQ-est}). "}
par(mfrow = c(1, 3))
rml(n = 500, tail = 0.8) %>% mlqqplot(tail = 0.8, log = 'xy', main = "Mittag-Leffler data")
flares_arrivals_MLparam <- (flares_arrivals %>% mlmle())$par
mlqqplot(flares_arrivals, 
         tail = flares_arrivals_MLparam["tail"], 
         scale = flares_arrivals_MLparam["scale"],
         log = 'xy', 
         main = "Threshold crossing times")
flares %>% ctrm() %>% thin(k = 500) %>% coredata() %>% 
  mlqqplot(tail = 0.8, log = 'xy', main = "Magnitudes")
```

@@

The estimates $\hat \beta$ and $\hat \sigma_0$ give an estimate of the 
distribution of exceedance times, dependent on the threshold $\ell$:
\begin{align*}
T(\ell) \sim {\rm ML}(\hat \beta, k^{-1/\hat \beta} \hat \sigma_0).
\end{align*}
For quick estimates of the Mittag-Leffler parameters we have used 
the method of log-transformed moments by @Cahoy2013.

@@

```{r flare-diagnostics-2, fig.height=7, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(2,2))
empcopula(flares_ctrm)
mlqqplot(flares_ctrm, log = 'xy', tail = 0.8, main = "Mittag-Leffler QQ Plot")
```

@@

```{r simulated-example, message=FALSE, fig.cap=caption}
k_max <- 500 
thin_sim_ctrm <- thin(ctrm = sim_ctrm, k = k_max)
par(mfrow=c(1,2))
MLestimates(ctrm = thin_sim_ctrm, tail = tail, scale = n^(1/tail))
```

>>>>

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
For all statistical computations we have used R [@R]. 
Code used for the analysis in this article has been organized into 
an R package `CTRM` (<https://github.com/UNSW-MATH/CTRM>).
The source code for this manuscript is available online at <https://github.com/UNSW-MATH/bursty-POT>.

@@

```{r thresholdedBursty, message=FALSE, fig.height=7, fig.width=9, fig.cap="\\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line)."}
#rpareto <- function(n,tail){
#          if(tail<=0) stop("tail must be positive")
#          rp <- (runif(n)/gamma(1-tail))^(-1/tail)
#          rp}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
#times<-cumsum(rpareto(n,tail))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctrm <- ctrm(data.frame(times, magnitudes)) 
par(mfrow = c(2,1))
plot(sim_ctrm, p = 0.01, main = "Simulated MRP") 
flares %>% ctrm() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") 
```

@@

\noindent For a scale parameter $\sigma > 0$, we write 
${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. 
The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, from the point of view of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11, and for algorithms, see e.g. the R package `MittagLeffleR`
[@MittagLeffleR].

@@

```{r fig:QQ-plots, fig.cap="Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\\label{fig:QQ-plots}", fig.height=5.5, fig.width=7}
par(mfrow = c(2,3))

flare_mags <- flares %>% ctrm() %>% thin(k=500) %>% coredata()
flares_exponent <- flare_mags %>% qqestplot_static()

flare_mags %>% tea::qqestplot(conf.int = TRUE)
abline(h = 0.8, lty = 2, col = 2)

flare_mags %>% mlqqplot(tail = 0.8, log = 'xy')
(flare_mags %>% MittagLeffleR::logMomentEstimator())[1:2] -> MLparams
flare_mags %>% qqline(distribution = function(p) qml(p, tail = 0.8, scale = MLparams[2]))

ML_data <- MittagLeffleR::rml(500, 0.8)
ML_exponent <- ML_data %>% qqestplot_static()

ML_exponent2 <- ML_data %>% tea::qqestplot(conf.int = TRUE)
abline(h=0.8, lty = 2, col = 2)

ML_data %>% mlqqplot(0.8, log = 'xy')
qqline(ML_data, distribution = function(p) qml(p, 0.8), col = 2)

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis). The reciprocal slope is returned as the estimate of the tail 
exponent.
For instance, the top-left panel in Figure \ref{fig:QQ-plots} shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`. The dynamic QQ-estimator plot (top-center panel)
plots the tail exponent estimate for the largest values at cutoffs 
varying up to the 5th order statistic, and the region of stability at 
0.8 indicates a recommended estimate.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill or QQ-Plot estimator becomes. 
The bottom-left panel of Figure \ref{fig:QQ-plots}, for instance, shows the 
static QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). 
Even by looking at different cutoffs via the dynamic QQ-estimator plot 
(bottom-center panel) one hardly identifies $0.8$ as a stable value, and 
thus as a clear candidate for a tail parameter estimate.

QQ-estimators, and the closely related Hill estimator, are hence not 
suitable to detect a heavy-tailed Mittag-Leffler distribution. Indeed, 
QQ-estimator plots of exponentially distributed data are virtually 
indistinguishable from the bottom-left panel. This shows that 
_Mittag-Leffler distributed data may look like exponentially distributed 
data if examined via a QQ-estimator_.

@@

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly 
(see Figure \ref{fig:QQ-plots}, right column). 
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in `MittagLeffleR` 
[@MittagLeffleR]. 
Since the exponential distribution is nested in the Mittag-Leffler 
family of distributions, a standard likelihood ratio test can be 
performed, with the exponential distribution as a null model against 
a Mittag-Leffler distribution as the alternative. 
As an example, the threshold crossing times for the solar flare dataset
yield a difference in deviance of 
$\approx$ `r round(2 * (mlmle_out$loglik - mlmle_out$null_loglik))`, 
which evaluates to a strongly significant $\chi^2_1$ test statistic 
(p-value = $10^{`r round(log10(mlmle_out$p_value))`}$) for the null
hypothesis $\beta = 1$.

@@

The estimates $\hat \beta$ and $\hat \sigma_0$ give an estimate of the 
distribution of exceedance times, dependent on the threshold $\ell$:
\begin{align*}
T(\ell) \sim {\rm ML}(\hat \beta, k^{-1/\hat \beta} \hat \sigma_0).
\end{align*}
For quick estimates of the Mittag-Leffler parameters we have used 
the method of log-transformed moments by @Cahoy2013.
We have verified the validity of our estimation algorithm  via simulations, 
see the appendix.

@@

```{r flare-diagnostics-2, fig.height=7, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(2,2))
empcopula(flares_ctrm)
flares_ctrm %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.8, main = "Mittag-Leffler QQ Plot")
```

@@

```{r simulated-example, message=FALSE, fig.cap=caption}
k_max <- 500 
thin_sim_ctrm <- thin(ctrm = sim_ctrm, k = k_max)
par(mfrow=c(1,2))
MLestimates(ctrm = thin_sim_ctrm, tail = params$tail, scale = n^(1/params$tail))
```
